---
title: "Advanced usage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(weasel)
long_data <- generate_weasel_dummy_data(
  n_ids = 400,
  n_times = 12,
  seed = 7
)
```

## Gap semantics

Two kinds of absence are treated differently throughout the package:

* **Endpoints**: whether a respondent is observed at the first and last
  wave of the window. Controlled by `require_endpoints` in scenarios
  and reported as `has_lower` / `has_upper` in the id metrics.
* **Interior gaps**: runs of missing waves strictly between a
  respondent's first and last observed wave within the window. A
  respondent observed at waves 2 to 7 of a 1-to-8 window has no
  interior gap; late entry and early exit are not gaps.

This separation keeps the constraints orthogonal: you can demand
anchored endpoints while tolerating interior gaps, or the reverse.

## Custom scenarios

`weasel_plan()` accepts a custom scenario table. It must contain the
columns `scenario`, `require_endpoints`, `max_missing`, `n_gap_max`,
and `max_gap_max`; the table is validated and a missing column raises
an immediate error rather than silently selecting nobody.

```{r}
my_scenarios <- data.frame(
  scenario          = c("complete_case", "one_gap_ok", "very_lenient"),
  require_endpoints = c(TRUE, TRUE, FALSE),
  max_missing       = c(0, 2, 4),
  n_gap_max         = c(0, 1, 4),
  max_gap_len       = c(0, 2, 4)
)

plan_custom <- weasel_plan(
  long_data,
  id = "id",
  wave = "time",
  span = "core",
  core_len = 8,
  scenarios = my_scenarios
)

weasel_print_table(
  weasel_compare_scenarios(plan_custom),
  title = "Custom scenarios", digits = 3
)
```

## Scoring and weights

The composite score is built only from observed subset quality. With
default weights it is:

```
score = 2.0 * mean_prop_present      (coverage)
      + 1.2 * endpoint_rate          (endpoints)
      + 0.8 * n_ids / max(n_ids)     (relative sample size)
      - 0.6 * worst_missing / L      (worst retained respondent)
      - 0.4 * (mean_n_gap + mean_max_gap) / L   (gap structure)
```

Scenarios that retain no respondents receive `NA` and are never
recommended; ties go to the larger sample. You can reweight any term:

```{r}
size_first <- weasel_compare_scenarios(
  plan_custom,
  weights = c(size = 3, coverage = 1)
)
size_first$scenario[size_first$recommended]
```

## Non-consecutive wave schedules

Both pipelines evaluate presence over a wave grid. The default,
`grid = "consecutive"`, treats every integer between the span bounds as
a scheduled wave. That is wrong for biennial panels or panels whose
waves are recorded as years: every skipped year would count as a wave
missed by everyone.

```{r}
biennial <- generate_weasel_dummy_data(
  n_ids = 200,
  waves = seq(2008, 2022, by = 2),
  seed = 11
)
sort(unique(biennial$time))
```

With the consecutive default, weasel now warns (a
`weasel_empty_waves` warning) and every scenario collapses:

```{r}
p_wrong <- weasel_plan(biennial, "id", "time", span = "full")
weasel_compare_scenarios(p_wrong)$n_ids
```

`grid = "observed"` evaluates presence on the schedule that actually
occurs in the data. Interior gaps then mean skipped *scheduled* waves,
and the span length `L` counts observed waves:

```{r}
p_biennial <- weasel_plan(biennial, "id", "time",
                          span = "full", grid = "observed")
p_biennial
```

`span = "core"` combines with `grid = "observed"` as you would expect:
the window is a block of `core_len` consecutive *scheduled* waves.
The same `grid` argument exists in `set_weasel_scope()` for the
exploratory pipeline.

## Sensitivity of the sample size to the tolerances

How strongly does the retained sample react to the tolerance settings?
`weasel_sensitivity()` sweeps every combination of the supplied values
against the per-respondent metrics already stored in the plan, so it is
cheap even for large panels:

```{r}
sens <- weasel_sensitivity(
  plan_custom,
  require_endpoints = c(TRUE, FALSE),
  max_missing = 0:2,
  n_gap_max = 0:1,
  max_gap_len = 0:2
)
weasel_print_table(sens, title = "Tolerance sweep", n = 10)
```

A steep jump in `n_ids` between adjacent rows tells you exactly which
constraint is binding; reviewers tend to ask for this table.

## Selectivity: who gets excluded?

Selecting on participation completeness can bias a sample if retention
relates to the covariates you study. `weasel_selectivity()` compares
retained respondents with those observed in the span but excluded,
using each respondent's value at their first observed wave (or their
within-span mean with `at = "mean"`), and reports standardized mean
differences:

```{r}
sel <- weasel_selectivity(plan_custom, "one_gap_ok")
weasel_print_table(sel, title = "Retained vs excluded", digits = 3)
```

Absolute SMD values around 0.1 or larger are commonly read as
noteworthy imbalance; the synthetic data here has no such mechanism,
so the differences stay small.

## Plans without attached data

By default the plan object carries the original data, which is
convenient but heavy for large panels (for example when saving plans
with `saveRDS()`). Use `keep_data = FALSE` and pass the data back in
where needed:

```{r}
p_light <- weasel_plan(long_data, "id", "time",
                       span = "full", keep_data = FALSE)
nrow(weasel_apply(p_light, "lenient", data = long_data))
```

`weasel_sensitivity()` and `weasel_justify_subset()` keep working on
such plans because they only need the stored metrics.

## Span selection in detail

`span = "core"` slides a window of `core_len` consecutive grid waves
over the full observed range and scores each window by mean per-wave
respondent coverage; with `grid = "consecutive"`, waves with no
observations contribute zero, so the chosen window is always a
contiguous block of `core_len` integer waves. Use `span = "full"` to
keep the entire range:

```{r}
plan_full <- weasel_plan(long_data, "id", "time", span = "full")
c(lower = plan_full$lower, upper = plan_full$upper)
```

## Auditing per-respondent metrics

The plan object exposes the per-respondent table that the scenarios
filter on, which is useful for custom diagnostics:

```{r}
m <- plan_custom$id_metrics
weasel_print_table(head(m, 8), title = "Per-respondent metrics")

# distribution of interior gap counts
table(m$n_gap)
```

## Scope pipeline constraints

The exploratory pipeline accepts the same structural vocabulary as the
scenario tables. `min_present` sets the minimum number of observed
waves a respondent needs to be kept (default 1, so exploration shows
everyone); `max_missing`, `max_gap_len`, `n_gap_max`, and
`require_endpoints` apply the same constraints a plan scenario would:

```{r}
set_weasel_scope(
  long_data,
  id = "id", wave = "time",
  lower = 2, upper = 11,
  min_present = 6,   # at least 6 observed waves within the span
  max_gap_len = 1,   # no interior gap longer than 1 wave
  n_gap_max = 2      # at most 2 interior gaps
)
evaluate_weasel_scope()
weasel_reshape_to_wide()
weasel_summarize_waves()
weasel_print_table(weasel_filter_wave_summary(),
                   title = "Constrained patterns", n = 8)
weasel_clear_scope()
```

## Justification styles

`weasel_justify_subset()` writes the methods paragraph in three
registers:

```{r}
cat(weasel_justify_subset(plan_custom, "one_gap_ok", style = "concise"),
    "\n\n")
cat(weasel_justify_subset(plan_custom, "one_gap_ok", style = "methods",
                          author = "van den Bosch", year = "2026"),
    "\n")
```

The `extended` style adds sensitivity framing for reviewers:

```{r}
cat(weasel_justify_subset(plan_custom, "one_gap_ok", style = "extended"),
    "\n")
```

## Quiet mode and reproducibility

All status messages respect a single option:

```{r}
options(weasel.verbose = FALSE)
p_quiet <- weasel_plan(long_data, "id", "time", span = "full")
options(weasel.verbose = TRUE)
```

The data generator and the cosmetic console colours preserve your RNG
state, so calling weasel functions never shifts the random numbers your
own analysis draws afterwards:

```{r}
set.seed(99)
before <- rnorm(2)

set.seed(99)
tmp <- generate_weasel_dummy_data(n_ids = 20, n_times = 5, seed = 1)
after <- rnorm(2)

identical(before, after)
```
