---
title: "Introduction to weasel"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to weasel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## What weasel does

Longitudinal panel studies rarely deliver the data you wish you had.
Respondents skip waves, drop out, or return after long absences. Before
any substantive analysis you must decide which respondents and which
waves to keep, and that decision shapes everything downstream.

The weasel package (Wave-based Extraction and Selection for
Longitudinal Data) turns this selection step into an explicit,
inspectable, and reproducible workflow. Throughout the package, a
respondent counts as observed at a wave when a row with that (id, wave)
pair exists in the long-format data; a missed wave is simply an absent
row, which is how real panel files are usually delivered.

It offers two complementary pipelines:

1. A **scope pipeline** for interactive exploration of wave patterns.
2. A **plan pipeline** for comparing predefined selection scenarios and
   generating methods-section text.

```{r setup}
library(weasel)
```

## Synthetic example data

We start with the built-in generator, which layers several realistic
participation mechanisms (random skips, attention decay, permanent
attrition, and block dropout) and removes the corresponding rows:

```{r}
long_data <- generate_weasel_dummy_data(
  n_ids = 300,
  n_times = 12,
  seed = 2026
)
str(long_data)
```

Note that the grid is incomplete by design:

```{r}
nrow(long_data)
300 * 12
```

A small amount of item nonresponse (`NA` values inside outcome columns
of observed rows) is also injected via `prop_item_missing`; it is
distinct from wave-level missingness and does not affect selection.

## The scope pipeline

The scope pipeline holds its state inside the package itself, not in
your global environment, so it never interferes with your own objects.

```{r}
set_weasel_scope(
  long_data,
  id = "id",
  wave = "time",
  lower = 2,
  upper = 11
)
evaluate_weasel_scope()
weasel_reshape_to_wide()
weasel_summarize_waves()
```

At any point, `weasel_scope_info()` shows what the scope contains and
how far the pipeline has progressed:

```{r}
weasel_scope_info()
```

The pattern summary groups respondents by their participation pattern.
In the table, `waves` shows the pattern (a `.` marks a missing wave),
`n` is the number of observed waves in that pattern, and `ids` counts
the respondents who share it:

```{r}
weasel_print_table(weasel_filter_wave_summary(),
                   title = "Wave patterns", n = 10)
```

`weasel_filter_wave_summary()` narrows the table. For instance,
patterns with at least 9 observed waves that are shared by at least 5
respondents:

```{r}
weasel_print_table(
  weasel_filter_wave_summary(n_range = c(9, 10), ids_range = c(5, Inf)),
  title = "Dense, common patterns"
)
```

`weasel_get_data_by_row()` then pulls the original long-format rows
behind any pattern row (or several rows at once), ready for analysis:

```{r}
subset_row1 <- weasel_get_data_by_row(1)
weasel_print_table(head(subset_row1, 8), title = "Most common pattern, data")
```

You can also constrain patterns structurally when setting the scope.
The `gap` argument limits the length of interior gaps (runs of missing
waves strictly between a respondent's first and last observed wave),
and `n_gap` limits how many such gaps are allowed:

```{r}
set_weasel_scope(long_data, "id", "time", gap = 1, n_gap = 1)
evaluate_weasel_scope()
weasel_reshape_to_wide()
weasel_summarize_waves()
weasel_print_table(weasel_filter_wave_summary(),
                   title = "Gap-constrained", n = 5)
```

When you are done, clear the scope:

```{r}
weasel_clear_scope()
```

## The plan pipeline

For confirmatory work you usually want a small number of named,
defensible scenarios rather than free exploration. `weasel_plan()`
evaluates three defaults:

* `anchored_strict`: observed endpoints, no missing waves.
* `anchored_balanced`: observed endpoints, at most one missing wave and
  one interior gap.
* `lenient_info_max`: no endpoint requirement, up to two missing waves
  and two interior gaps.

```{r}
plan_obj <- weasel_plan(long_data, id = "id", wave = "time", span = "core")
plan_obj
```

With `span = "core"` the package slides a window of `core_len`
consecutive waves across the full wave range and picks the window with
the highest average per-wave coverage.

```{r}
comparison <- weasel_compare_scenarios(plan_obj)
weasel_print_table(comparison, title = "Scenario comparison", digits = 3)
```

The `score` column is computed from observed subset quality (coverage,
endpoint rate, relative sample size, worst-case missingness, and gap
structure), never from a scenario's configured tolerances; see
`?weasel_compare_scenarios` for the exact formula and how to override
the weights. A sentence version is available too:

```{r}
cat(weasel_compare_to_sentence(comparison), "\n")
```

Audit a scenario before committing to it:

```{r}
s <- weasel_summarize_subset(plan_obj, "anchored_balanced")
weasel_print_table(s$headline, title = "Headline", digits = 3)
weasel_print_table(s$per_wave_coverage, title = "Per-wave coverage")
cat(weasel_subset_to_sentence(s), "\n")
```

Finally, extract the analysis-ready subset:

```{r}
analysis_data <- weasel_apply(plan_obj, "anchored_balanced")
dim(analysis_data)
```

## Justification text for your methods section

Documenting the selection is as important as making it. The package
drafts that paragraph for you:

```{r}
cat(weasel_justify_subset(plan_obj, "anchored_balanced"), "\n")
```

See the "Advanced usage" vignette for custom scenarios, scoring
weights, non-consecutive wave schedules (biennial panels), the
sensitivity and selectivity diagnostics, and writing style options.
